\sys: Controllable Program Regeneration

### 2025-07-17 docs/plos/draft.tex
Update intro and examples
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

\begin{abstract}
Modern software systems are constantly evolving—requiring developers to
  refactor legacy code, adapt to changing APIs, translate across languages, and
  patch security vulnerabilities. These transformations are often brittle,
  manual, and error-prone. Large language models (LLMs) offer new opportunities
  for automating such tasks, but their outputs are shaped by surface cues, lack
  semantic guarantees, and require brittle prompting strategies. This paper
  presents \sys, a system for controllable program regeneration that combines
  logic-based planning with LLM-guided synthesis.
Kvasir operates in four stages: (1) a symbolic planner constructs a
transformation plan using logic rules and a domain-aware knowledge base; (2)
property extractors analyze the input program to gather relevant semantic
information; (3) a language model synthesizes candidate programs guided by
a structured prompt assembled from extracted properties; and (4) a verification
and feedback loop checks whether the output meets the desired properties and
iteratively refines the generation process through additional feedback.
\sys is evaluated on a diverse set of program transformation tasks
including security hardening, language translation, idiomatic rewriting, de-obfuscation,
and modular refactoring.
Target programs are sourced from high-profile supply-chain attacks, 
online repositories, programming competitions, and real-world software projects.
\sys produces high-quality outputs that satisfy goals in all but two cases,
even on instances where a state-of-the-art LLM fails to do so.
\end{abstract}
\maketitle

@@

\begin{figure}[t]
  % https://docs.google.com/drawings/d/1ozMu2B0RUhyLWC_rM6GLBrioov80-M06CBqYRAhCk18/edit
  \includegraphics[width=.9\columnwidth]{figs/kvasir_projection.pdf}
  \caption{\textbf{\sys overview}
Given a transformation query, and a source program, \sys extracts a minimal set
  of properties,
  guided by a logic engine and knowledge base.
  It then synthesizes a new program $P'$ that satisfies the query, verifying
  that wanted properties are present and unwanted properties are absent in $P'$.
}
  \label{fig:projection}
\end{figure}

@@

Recent advances in large language models (LLMs) have opened new possibilities for automating program transformation and synthesis tasks, making them attractive tools for software evolution, albeit with caveats.
This flexibility has positioned them as attractive tools for software evolution
compared to traditional program synthesis techniques, which often struggle
to scale to real-world settings or across languages and require significant developer involvement~\cite{reynolds2019syguscomp,leino2016dafny,wu2023programming,dynamoth2016,cambronero2019active}.
LLMs can translate code across languages~\cite{ou2025enhancingllmbasedcodetranslation},
refactor complex modules~\cite{ziftci2025migrating},
and even generate entire components from scratch~\cite{huynh2025largelanguagemodelscode}.
However, surface-level patterns in the input shape their outputs~\cite{yang2025evaluatinggeneralizationcapabilitieslarge},
they struggle with negative constraints~\cite{hwang2024thinkpinkelephant,jiang2024llmsdreamelephantswhen},
often overfit to spurious input cues~\cite{xu2023llmfoolitselfpromptbased, wu2023deceptpromptexploitingllmdrivencode},
and lack mechanisms for enforcing precise behavioral properties on the generated code~\cite{roh2025breakthechainreasoningfailuresllms}.
As a result, developers must prompt and steer them through brittle heuristics or manual trial-and-error, leading to workflows that are difficult to audit, integrate, or systematically reason about.
Bridging this gap calls for a declarative interface---one that allows developers to specify explicit semantic constraints and transformation goals, and to reason about their satisfaction in a principled way.

This paper present \sys, a system for \emph{property-guided program regeneration}.
\sys allows users
to express transformation goals declaratively, as logical constraints over
program properties such as input/output behavior, side-effect freedom, target
language, trace equivalence, or modular structure.
Given a source program and
a query, \sys extracts a minimal set of relevant properties, guided by a
logic engine and a domain-specific knowledge base, and synthesizes a new
program that satisfies the desired constraints.

Three key insights shape the design of \sys.
First, modern software is highly modular, making it feasible to reason about transformations at the granularity of individual components or functions.
Second, the program regeneration setting provides a unique advantage: the original program serves as a concrete ground-truth, enabling semantic comparisons and verification of the transformed output, something impossible in traditional synthesis tasks.
Third, the underlying synthesis technology---particularly LLMs---is evolving rapidly; a declarative interface that harnesses this progress without disrupting the framework's end user.

Guided by these insights, \sys casts transformation as a constrained synthesis problem.
At its core is a property-driven pipeline: users express goals as logical constraints over semantic properties, such as input/output behavior, side-effect freedom, target language, or structural modularity.
\sys extracts a set of relevant properties, derived from its logic engine and a domain-aware knowledge base, and synthesizes a new program that satisfies the specified constraints.
If no such program exists, \sys returns a failure explanation from the unsatisfiable core.

\heading{Deployment scenarios and limitations}
\sys supports a range of deployment scenarios.
It is particularly well-suited for use during maintenance and auditing phases, where developers seek to refactor or replace components, especially when the original code is difficult to understand, modify, or is unavailable and the component exists only as a binary.
It can also assist during development, for instance when integrating third-party modules that must be adapted to local conventions or hardened against security vulnerabilities.
However, \sys is not designed for whole-program synthesis or large-scale migrations.
Its strength lies in scoped regeneration tasks, typically at the granularity of a function, method, or small module---where transformation goals can be explicitly stated and verified in isolation.
To this end, \sys functions not only as a standalone tool but also as a
library, where other components can invoke it as a programmable,
constraint-guided regeneration module.

@@

@@

We illustrate the capabilities of \sys by applying it to a representative game
application.
This example is selected to highlight the
diversity of transformation goals that arise even within a single codebase:
security hardening, cross-language migration, modernization of unsafe idioms,
and structural refactoring.

In each case, \sys first selects a minimal set of properties to extract from
the original program, guided by the transformation query and the
domain-specific knowledge base.
It then synthesizes a new program that
satisfies the declarative constraints while preserving the original
program's behavior to the extend expressible by the extracted properties.

@@

\heading{An unidiomatic C function in a physics library}
The fast inverse square root routine, originally developed for the Quake III
engine~\cite{fast_inv_sqrt}, is included in the game's physics and vector math
library as an optimization for computing normalized vectors.
The implementation
exploits type punning to manipulate IEEE floats at the bit level—a technique
that relies on undefined behavior and reduces portability:
\begin{listing}
\begin{minted}{c}
float Q_rsqrt(float number) {
 long i; float x2, y;
 const float threehalfs = 1.5F;
 x2 = number * 0.5F; y  = number;
 i  = * ( long * ) &y; // evil bit level hack
 i  = 0x5f3759df - ( i >> 1 );
 y  = * ( float * ) &i;
 y  = y * ( threehalfs - ( x2 * y * y ) );
 return y; }
\end{minted}
\caption{An unidiomatic C implementation of the fast inverse square root function popularized by the game Quake III~\cite{fast_inv_sqrt}.}
\end{listing}
The following query for \sys uses lines of code as a coarse metric of
idiomaticity. \sys regenerates the program, minimizing this metric while
preserving observable behavior over a representative set of numeric inputs:
\begin{wrapfigure}[3]{r}{.7\columnwidth}
  \vspace{-5pt}
  \begin{minted}{prolog}
graph(p_, I, O) :- graph(p, I, O).
#min N: len(p_, N).
  \end{minted}
\end{wrapfigure}
\sys extracts the set of properties \texttt{signature(F)} and \texttt{graph(F)}
from the original function, where \texttt{signature(F)} is the function's
signature (including its name, return type, and arguments) and
\texttt{graph(F)} captures the I/O behavior on a generated set of
floating-point inputs.
In this case, where the goal is to optimize an aspect of the program, \sys
re-prompts the model to regenerate the program until it can make no further
improvements or observes a negligible difference after two consecutive
iterations.

@@

\heading{A monolithic gameplay function}
Consider a JavaScript function in the game's scripting environment that
retrieves and displays a player's saved inventory from a local SQLite database.
The function combines database initialization and query execution in a single
block of logic:

\begin{minted}{javascript}
function getInventory(playerId) {
const db = new Database("game.db");
const rows = db.prepare("SELECT
             item FROM inventory
             WHERE player_id = ?").all(playerId);
return rows.map(row => row.item); }
\end{minted}

@@

\begin{wrapfigure}[6]{r}{.66\columnwidth}
  \begin{minted}{prolog}
db('game.db').func(f1).func(f2).
sql_trc(p_,I,T):-sql_trc(p,I,T),
                 graph(p,I,_).
graph(call(f2,call(f1),I),I,O):-
                   graph(p,I,O).
  \end{minted}
\end{wrapfigure}

@@

\heading{Key results}
\sys is able to effectively transform all programs across the above tasks,
producing high-quality outputs that satisfy user-specified constraints.
A total
of six plugins (\ttt{input}, \ttt{graph}, \ttt{signature}, \ttt{pure}, \ttt{trace\_sql}, and \ttt{javascript2haskell})
were required to perform the above transformations, each
contributing either a set of extracted and verified properties, or a
transformation, and a set of pre- and post-conditions that the logic engine
uses to construct the regeneration plan.

@@

\heading{Logic programming and constraint solving}
Logic programming has long served as a foundation for reasoning in programming
tools, offering a declarative model for expressing properties and constraints.
Answer Set Programming (ASP), in particular, provides a rich framework for
expressing defaults, exceptions, and optimization via stable model semantics
\cite{Gelfond_2000, Gelfond_2002, Eiter_2009}. 
ASP has been used for
declarative program analysis \cite{benton2007interactive}, automated planning
\cite{nguyen2020explainable, son2022answersetplanningsurvey}, and knowledge
representation in verification and synthesis pipelines.

@@

\heading{Targeted regeneration}
\sys is best used as a surgical tool for targeted regeneration tasks, not for whole-program synthesis or full language migration.
It operates most effectively on localized units of code---such as individual functions, methods, or small modules, where behavior is self-contained and transformation goals can be expressed through explicit constraints.
% Typical applications include rewriting high-risk, deprecated, or poorly understood code segments where correctness is critical and semantic preservation is essential.
Broad use may yield incomplete results and can be bottlenecked by the limited context of the back-end model \sys uses.

>>>>

\begin{abstract}
  Software evolution tasks---like security hardening, porting critical components to safer 
  languages, or refactoring opaque legacy code---are time-consuming and brittle.
  Large language models (LLMs) have shown promise in automating such tasks,
  but their outputs are shaped by surface cues, lack
  semantic guarantees, and their usage often boil down to brittle prompting strategies.
This paper presents \sys, a system for controllable program regeneration that
  combines declarative property constraints with LLM-guided synthesis. Unlike
  prior approaches, \sys treats the input program as an untrusted artifact and
  explicitly projects it into a minimal set of verifiable properties, such as
  input/output behavior, execution traces, and API signatures. These properties
  serve as constraints during synthesis and are checked post hoc to ensure
  correctness.
\sys is evaluated on a diverse set of program transformation tasks
including security hardening, language translation, idiomatic rewriting, de-obfuscation,
and modular refactoring showing strong results, even on tasks where state-of-the-art LLMs fail.
\end{abstract}

\maketitle

@@

\begin{figure}[t]
  % https://docs.google.com/drawings/d/1ozMu2B0RUhyLWC_rM6GLBrioov80-M06CBqYRAhCk18/edit
  \includegraphics[width=.9\columnwidth]{figs/kvasir_projection.pdf}
  \caption{\textbf{Property projection and verifiable regeneration in \sys.}
  Instead of transforming source code directly, \sys projects the original
  program into a minimal set of verifiable properties---such as input/output
  examples, system call traces, \etc.
  These properties serve as declarative
  constraints for synthesis.
  The regenerated implementation is then checked to
  ensure it satisfies the specified properties, providing a verifiable
  foundation for trust unlike purely prompt-based LLM workflows.}
  \label{fig:projection}
\end{figure}

@@

Recent advances in natural-language processing, particularly large language
models (LLMs) have opened new possibilities for automating program
transformation and synthesis tasks,
making them attractive tools for software
evolution, albeit with caveats.
This flexibility has positioned them as attractive tools for software evolution
compared to traditional program synthesis techniques, which often struggle
to scale to real-world settings or across languages and require significant developer involvement~\cite{reynolds2019syguscomp,leino2016dafny,wu2023programming,dynamoth2016,cambronero2019active}.
LLMs can translate code across languages~\cite{ou2025enhancingllmbasedcodetranslation},
refactor complex modules~\cite{ziftci2025migrating},
and even generate entire components from scratch~\cite{huynh2025largelanguagemodelscode}.
However, surface-level patterns in the input shape their outputs~\cite{yang2025evaluatinggeneralizationcapabilitieslarge},
they struggle with negative constraints~\cite{hwang2024thinkpinkelephant,jiang2024llmsdreamelephantswhen},
often overfit to spurious input cues~\cite{xu2023llmfoolitselfpromptbased, wu2023deceptpromptexploitingllmdrivencode},
and lack mechanisms for enforcing precise behavioral properties on the generated code~\cite{roh2025breakthechainreasoningfailuresllms}.
As a result, developers must prompt and steer them through brittle heuristics or manual trial-and-error, leading to workflows that are difficult to audit, integrate, or systematically reason about.
Bridging this gap calls for an that allows developers to specify explicit semantic constraints and transformation goals, and to reason about their satisfaction in a principled way.

This paper explores the feasibility of combining declarative property
constraints with LLM-based synthesis to support \emph{property-guided program regeneration}.
The key idea is to treat the original program as an untrusted reference from
which to extract minimal, verifiable properties (\eg I/O behavior,
function signatures, program traces), and to synthesize a new implementation that satisfies those
constraints without overfitting to potentially spurious details of the original source code.
The engine for this experimentation is the \sys prototype.
\sys allows users
to express transformation goals declaratively, as logical constraints over
program properties.
At a high level, given a source program and a query, \sys uses the provided
logical constraints and its internal knowledge base to extract a set of
feasible and relevant program properties.
It then prompts an LLM to synthesize
candidate programs, which are subsequently verified against the desired properties.
This approach treats program transformation as projecting the original program over key properties.
Intuitively, regeneration then leverages LLM's high-dimensional internal representation~\cite{jin2024emergent,tao2024llms,huh2024platonicrepresentationhypothesis}, 
which lifts the projection back to a concrete implementation~(\cref{fig:projection}).
Unlike prompt-based approaches that require
extensive curation of examples~\cite{dilhara2024unprecedented, khattab2023dspy, cummins2024donttransformcodecode},
\sys treats the input program as an
untrusted artifact and explicitly projects only minimal, verifiable properties
into the regeneration process.

Three key insights shape the design of \sys.
First, modern software is highly decomposable~\cite{vfunction2024modular, isoline2018decomposition, schechter2011visualizing, breakapp:ndss:2018}, making it feasible to reason about transformations at the granularity of individual modules, libraries, or functions.
Second, the program \emph{regeneration} setting provides a unique advantage: the original program serves as a concrete ground-truth, enabling semantic comparisons and verification of the transformed output, something impossible in traditional synthesis tasks.
Third, the underlying synthesis technology---particularly LLMs---is evolving rapidly; a declarative interface harnesses this progress without disrupting the framework's end user.

% Guided by these insights, \sys casts transformation as a constrained synthesis problem.
% At its core is a property-driven pipeline: users express goals as logical constraints over semantic properties, such as input/output behavior, side-effect freedom, target language, or structural modularity.
% \sys extracts a set of relevant properties, derived from its logic engine and a domain-aware knowledge base, and synthesizes a new program that satisfies the specified constraints.
% If no such program exists, \sys returns a failure explanation from the unsatisfiable core.

\heading{Deployment scenarios and limitations}
\sys supports a range of deployment scenarios.
It is particularly well-suited for use during maintenance and auditing phases, where developers seek to refactor or replace components, especially when the original code is difficult to understand, modify, or is unavailable and the component exists only as a binary.
It can also assist during development, for instance when integrating
third-party modules that must be adapted to local conventions or hardened
against security vulnerabilities.
For example, a developer can regenerate a C library as idiomatic
JavaScript while preserving I/O equivalence, or decompose a monolithic function
into modular components suitable for sandboxing.
However, \sys is not designed for whole-program synthesis or large-scale migrations.
Its strength lies in scoped regeneration tasks, typically at the granularity of a function, method, or small module---where transformation goals can be explicitly stated and verified in isolation.
To this end, \sys can function not only as a standalone tool but also as a
library, where other components can invoke it as a programmable,
constraint-guided regeneration module.

@@

\heading{Contributions}
The paper's contributions are:
\begin{itemize}
 \item A property-aware transformation framework, combining logic-based reasoning, and LLM-assisted program synthesis~(\cref{sec:design});
 \item A declarative interface and accompanying DSL for program transformation, allowing users to express semantic goals as logical queries over properties of the regenerated program~(\cref{sec:dsl});
 \item A verification architecture that checks synthesized programs against the specified properties, providing feedback for refinement~(\cref{sec:verification});
 \item An empirical evaluation demonstrating \sys's feasibility across transformation scenarios~(\cref{sec:evaluation}).
\end{itemize}

@@

This examples illustrates the capabilities of \sys by applying it to a representative game
application.
This example is selected to highlight the
diversity of transformation goals that arise even within a single codebase:
security hardening, cross-language migration, modernization of unsafe idioms,
and structural refactoring.

@@

\heading{An unidiomatic C function in a physics library}
The fast inverse square root routine, originally developed for the Quake III
engine~\cite{fast_inv_sqrt}, is included in the game's physics and vector math
library as an optimization for computing normalized vectors.
The implementation
exploits type punning to manipulate IEEE floats at the bit level—a technique
that relies on undefined behavior and reduces portability:
\begin{listing}
\begin{minted}{c}
float Q_rsqrt(float number) {
 long i; float x2, y;
 const float threehalfs = 1.5F;
 x2 = number * 0.5F; y  = number;
 i  = * ( long * ) &y; // evil bit level hack
 i  = 0x5f3759df - ( i >> 1 );
 y  = * ( float * ) &i;
 y  = y * ( threehalfs - ( x2 * y * y ) );
 return y; }
\end{minted}
\caption{An unidiomatic C implementation of the fast inverse square root function popularized by the game Quake III~\cite{fast_inv_sqrt}.}
\end{listing}
The following query for \sys uses lines of code as a coarse metric of
idiomaticity. \sys regenerates the program, minimizing this metric while
preserving observable behavior over a representative set of numeric inputs:
\begin{wrapfigure}[3]{r}{.65\columnwidth}
  \vspace{-5pt}
  \begin{minted}{prolog}
graph(p_, I, O) :- graph(p, I, O).
#min N: len(p_, N).
  \end{minted}
\end{wrapfigure}
\sys extracts the set of properties \texttt{signature(F)} and \texttt{graph(F)}
from the original function, where \texttt{signature(F)} is the function's
signature (including its name, return type, and arguments) and
\texttt{graph(F)} captures the I/O behavior on a generated set of
floating-point inputs.
In this case, where the goal is to optimize an aspect of the program, \sys
re-prompts the model to regenerate the program until it can make no further
improvements or observes a negligible difference after two consecutive
iterations.

@@

\heading{A monolithic gameplay function}
Consider a JavaScript function in the game's scripting environment that
retrieves and displays a player's saved inventory from a local SQLite database~(\cref{listing:monolithic-sql}).
The function combines database initialization and query execution in a single
block of logic:

\begin{listing}
\begin{minted}{javascript}
function getInventory(playerId) {
const db = new Database("game.db");
const rows = db.prepare("SELECT
           item FROM inventory
           WHERE player_id = ?").all(playerId);
return rows.map(row => row.item); }
\end{minted}
\caption{A monolithic function that retrieves a player's inventory from a SQLite database.}
\label{listing:monolithic-sql}
\end{listing}

@@

\begin{wrapfigure}[7]{r}{.68\columnwidth}
  \begin{minted}{prolog}
db('game.db').func(f1).func(f2).
sql_trc(p_,I,T) :- sql_trc(p,I,T),
                   graph(p,I,_).
graph(call(f2,call(f1),I),I,O):-
              graph(p,I,O).
  \end{minted}
\end{wrapfigure}

@@

\heading{Key results}
\sys is able to effectively transform all programs across the above tasks,
producing high-quality outputs that satisfy user-specified constraints.
A total
of six property extraction/verification plugins (\ttt{graph}, \ttt{fsignt}, \ttt{sql\_trc}, and \ttt{llm\_translate})
were required to perform the above transformations, each
contributing either a set of extracted and verified properties, or a
transformation, and a set of pre- and post-conditions that the logic engine
uses to construct the regeneration plan.

@@

\heading{Logic programming and constraint solving}
Logic programming has long served as a foundation for reasoning in programming
tools, offering a declarative model for expressing properties and constraints.
Answer Set Programming (ASP), in particular, provides a rich framework for
expressing defaults, exceptions, and optimization via stable model semantics
\cite{Gelfond_2000, Gelfond_2002, Eiter_2009}. 
Paradigms like ASP have been used for
declarative program analysis \cite{benton2007interactive}, automated planning
\cite{nguyen2020explainable, son2022answersetplanningsurvey}, and knowledge
representation in verification and synthesis pipelines.

@@
